---
title: "Introduction to RNAseqDataDownloader"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to RNAseqDataDownloader}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(RNAseqDataDownloader)
```

## Overview

**RNAseqDataDownloader** is an R package that provides convenient functions to download RNA-seq and microarray data from major public repositories. The package streamlines the process of accessing data from:

- **GEO** (Gene Expression Omnibus) - NCBI's repository for gene expression data
- **TCGA** (The Cancer Genome Atlas) - Cancer genomics data
- **GTEx** (Genotype-Tissue Expression) - Gene expression across tissues
- **SRA** (Sequence Read Archive) - Raw sequencing data
- **HCA** (Human Cell Atlas) - Single-cell data
- **Single Cell Portal** - Single-cell RNA-seq studies

## Key Features

### 🚀 Fast and Efficient
- Uses optimized download methods (GSEMatrix for GEO = 10-100x faster)
- Automatic retry with exponential backoff
- Local caching to avoid re-downloads
- Configurable timeouts for large files

### 🧬 RNA-seq Specific
- **NEW**: Direct RNA-seq quantification data from NCBI GEO
- Pre-computed gene counts (HISAT2 + featureCounts)
- Returns SummarizedExperiment objects ready for DESeq2/edgeR

### 📊 Comprehensive Data Access
- Expression matrices
- Sample metadata
- Platform annotations with updated gene mappings
- Supplementary files (CEL, FASTQ, etc.)

### 🔧 Flexible and Robust
- Filter by platform, samples, or file types
- Handle multi-platform studies
- Comprehensive error handling
- Detailed progress reporting

## Installation

```{r installation, eval=FALSE}
# Install from GitHub
# install.packages("devtools")
devtools::install_github("vanhungtran/RNAseqDataDownloader")

# Load the package
library(RNAseqDataDownloader)
```

## Quick Start Examples

### Download from GEO

```{r geo-quick}
# Download a GEO Series (fastest method)
result <- download_geo("GSE12345")

# Download RNA-seq quantification data (NEW!)
result <- download_geo("GSE164073", data_type = "rnaseq")

# Download with updated gene annotations
result <- download_geo("GSE12345", AnnotGPL = TRUE)
```

### Download from TCGA

```{r tcga-quick}
# Download TCGA data
result <- download_tcga(
  project = "TCGA-BRCA",
  data_type = "Gene Expression Quantification",
  workflow_type = "HTSeq - Counts"
)
```

### Download from GTEx

```{r gtex-quick}
# Download GTEx data
result <- download_gtex(
  tissue = "Whole Blood",
  data_type = "gene_tpm"
)
```

## Available Databases

Check which databases are supported:

```{r databases}
available_databases()
```

## Main Functions

### GEO Downloads: `download_geo()`

Download data from Gene Expression Omnibus with extensive options:

```{r geo-functions}
# Basic download
download_geo("GSE12345")

# RNA-seq quantification (NEW!)
download_geo("GSE164073", data_type = "rnaseq")

# With filtering
download_geo("GSE12345", 
            data_type = "raw",
            filter_files = "\\.cel\\.gz$")

# Multi-platform filtering
download_geo("GSE12345", 
            platform = "GPL570")

# With updated annotations
download_geo("GSE12345", 
            AnnotGPL = TRUE)
```

**Key parameters:**
- `data_type`: "matrix" (fast), "raw", "both", "rnaseq", "full", "soft"
- `AnnotGPL`: Use updated gene annotations (TRUE/FALSE)
- `platform`: Filter specific platforms
- `filter_files`: Regex to filter supplementary files
- `timeout`: Download timeout in seconds
- `max_retries`: Number of retry attempts

See `vignette("geo_download")` for comprehensive examples.

### TCGA Downloads: `download_tcga()`

Download data from The Cancer Genome Atlas:

```{r tcga-functions}
# Download gene expression
download_tcga(
  project = "TCGA-BRCA",
  data_type = "Gene Expression Quantification"
)

# Download multiple data types
download_tcga(
  project = "TCGA-LUAD",
  data_type = c("Gene Expression", "Methylation")
)
```

See `vignette("tcga_download")` for detailed examples.

### GTEx Downloads: `download_gtex()`

Download data from Genotype-Tissue Expression project:

```{r gtex-functions}
# Download gene expression
download_gtex(
  tissue = "Brain",
  data_type = "gene_tpm"
)
```

### SRA Downloads: `download_sra()`

Download raw sequencing data from Sequence Read Archive:

```{r sra-functions}
# Download SRA data
download_sra(
  accession = "SRR123456"
)
```

### Single-Cell Data

```{r sc-functions}
# Human Cell Atlas
download_hca(
  project = "immune_cells"
)

# Single Cell Portal
download_sc_atlas(
  study = "melanoma"
)
```

## What's New (Latest Update)

### GEO Function Enhancements (October 2025)

Based on GEOquery 2.99.0+ from GitHub:

1. **RNA-seq Quantification Support** 🆕
   - Download pre-computed gene counts from NCBI
   - Returns SummarizedExperiment objects
   - Ready for differential expression analysis

2. **Annotation GPL** 🆕
   - Updated gene annotations with current Entrez mappings
   - Better feature annotations

3. **Enhanced Performance**
   - 10-100x faster with GSEMatrix format
   - Better handling of multi-platform studies
   - Improved memory efficiency

4. **Better Error Handling**
   - Exponential backoff retry logic
   - Automatic fallback mechanisms
   - More informative error messages

5. **Flexible Filtering**
   - Platform-specific downloads
   - Sample subsetting
   - File type filtering with regex

## Typical Workflows

### Workflow 1: Microarray Data Analysis

```{r workflow1}
# 1. Download data
result <- download_geo("GSE12345", 
                      data_type = "matrix",
                      AnnotGPL = TRUE)

# 2. Extract ExpressionSet
eset <- result$geo_object[[1]]

# 3. Get components
expr <- Biobase::exprs(eset)
pheno <- Biobase::pData(eset)

# 4. Analyze with limma, etc.
```

### Workflow 2: RNA-seq Analysis

```{r workflow2}
# 1. Download RNA-seq counts
result <- download_geo("GSE164073", data_type = "rnaseq")

# 2. Get SummarizedExperiment
se <- result$geo_object

# 3. Use with DESeq2
library(DESeq2)
dds <- DESeqDataSet(se, design = ~ condition)
dds <- DESeq(dds)
```

### Workflow 3: Cross-Database Analysis

```{r workflow3}
# Download from multiple sources
geo_data <- download_geo("GSE12345")
tcga_data <- download_tcga("TCGA-BRCA", 
                          data_type = "Gene Expression")
gtex_data <- download_gtex("Breast", 
                          data_type = "gene_tpm")

# Integrate and analyze
```

## Configuration

### Set Download Directory

```{r config}
# Set default download directory
set_download_dir("~/my_data")

# Get current download directory
get_download_dir()

# Set timeout for large files
set_timeout(600)  # 10 minutes
```

## Helper Functions

### Metadata Extraction

```{r helpers}
# Extract comprehensive metadata
metadata <- extract_geo_metadata(geo_object)

# Get SRA links from GEO samples
sra_links <- get_geo_sra_links("GSE12345")
```

### Data Conversion

```{r conversion}
# Convert to different formats
df <- convert_geo_format(geo_object, format = "data.frame")
se <- convert_geo_format(geo_object, format = "SummarizedExperiment")

# Save as CSV
convert_geo_format(geo_object, 
                  format = "csv", 
                  output_file = "expression.csv")
```

### Batch Operations

```{r batch}
# Download multiple accessions
accessions <- c("GSE12345", "GSE67890", "GSE11111")
results <- batch_download_geo(accessions, data_type = "matrix")
```

## Best Practices

1. **Use appropriate data types:**
   - `"matrix"` for fastest downloads (microarray)
   - `"rnaseq"` for RNA-seq quantification data
   - `"raw"` for supplementary files
   - `"both"` for complete data

2. **Filter when possible:**
   - Specify `platform` for multi-platform studies
   - Use `filter_files` to download only needed files
   - Select specific `samples` if analyzing subset

3. **Handle large files:**
   - Increase `timeout` for big datasets
   - Use `max_retries` for network reliability
   - Enable `verbose` to monitor progress

4. **Use caching:**
   - Files are cached locally by default
   - Set `force_download = FALSE` to reuse cached files
   - Saves bandwidth and time

5. **Check metadata first:**
   - Examine `result$metadata` before analysis
   - Verify sample information and platform details
   - Check for batch effects or confounders

## Performance Tips

- **GSEMatrix format**: 10-100x faster than SOFT
- **Platform filtering**: Reduces download size
- **Local caching**: Avoid re-downloading
- **Batch downloads**: Process multiple accessions efficiently

## Troubleshooting

### Common Issues

1. **Timeout errors**: Increase `timeout` parameter
2. **Network issues**: Increase `max_retries`
3. **Memory issues**: Download raw files instead of matrix
4. **Missing RNA-seq data**: Automatically falls back to matrix

### Getting Help

```{r help, eval=FALSE}
# Function documentation
?download_geo
?download_tcga

# Vignettes
vignette("geo_download")
vignette("tcga_download")

# Check package info
packageVersion("RNAseqDataDownloader")
```

## Additional Resources

- **GEO**: https://www.ncbi.nlm.nih.gov/geo/
- **TCGA**: https://portal.gdc.cancer.gov/
- **GTEx**: https://gtexportal.org/
- **GEOquery**: https://bioconductor.org/packages/GEOquery
- **Package GitHub**: https://github.com/vanhungtran/RNAseqDataDownloader

## Citation

If you use this package, please cite:

```{r citation, eval=FALSE}
citation("RNAseqDataDownloader")
```

## Session Info

```{r session, eval=TRUE}
sessionInfo()
```
